\systemEnding: Bootstrapping Domain-Specific Languages for Neurally-Guided Bayesian Program Learning

### 2019-04-14 archive.tex
updated latex to new local figure
a little more writing

\begin{figure}
  \includegraphics[width = \textwidth]{figures/dpl/list}
  \includegraphics[width = \textwidth]{figures/dpl/logo}
  \caption{}\label{exampleDSL}
\end{figure}

@@

Our model, called \systemEnding, works through a novel kind
of wake/sleep or `dream' learning~\cite{hinton1995wake}, iterating
through a wake cycle -- where it solves \textbf{tasks} by writing programs (Figure~\ref{threeCycles} top)
-- and a pair of sleep cycles, both of which are loosely biologically
inspired. The first sleep cycle, which we refer to as
\textbf{consolidation}, grows its library of code by replying
experiences from waking and consolidating them into new code
abstractions (Figure~\ref{threeCycles} left).  This cycle is inspired by the formation of abstractions
during sleep memory consolidation~\cite{DUDAI201520}.  The second
sleep cycle, which we refer to as \textbf{dreaming}, improves the
agents knowledge of how to write code by training a neural network to
help quickly search for programs. The neural net is trained on replayed
experiences as well as `fantasies', or samples, from the learned
library (Figure~\ref{threeCycles} right).  These two kinds of dreams are inspired by the distinct
episodic replay and hallucination components of dream
sleep~\cite{fosse2003dreaming}.

@@

\begin{array}{lr}
  \begin{array}{l}  p_x = \argmax_p \probability[x|p]\probability[p|\mathcal{D}^*,\theta^*]\text{, for each task $x\in X$}\end{array}
    &\text{\emph{Wake}}\\\\
\begin{array}{rl}  
\mathcal{D}^* = \argmax_\mathcal{D}&\int\probability\left[\mathcal{D},\phantom{^*}\theta \right]\prod_{x\in X}\sum_{\substack{p\text{ found during waking}}} \probability[x|p]\probability[p|\mathcal{D},\phantom{^*}\theta]\;\text{d}\theta\\
\theta^* = \argmax_\theta&\phantom{\int}\probability\left[\mathcal{D}^*,\theta \right]\prod_{x\in X}\sum_{p\text{ found during waking}} \probability[x|p]\probability[p|\mathcal{D}^*,\theta]
 \end{array}
&\text{\emph{Consolidation sleep}}\\\\
\begin{array}{l} Q(p|x)\approx \probability[p|x,\mathcal{D}^*,\theta^*]\end{array}
  &\text{\emph{Dream sleep}}
\end{array}

@@

\section{Related work}

We build on several generations of research in AI, program synthesis,
and cognitive science, with program induction being one of the oldest
theoretical frameworks for concept learning within artificial
intelligence~\cite{solomonoff1964formal}, and the conceptually allied
`Language of Thought Hypothesis' being almost as
old~\cite{fodor1975language}. Recent neural program synthesis systems
pair a fixed domain-specific programming language (a `DSL') to a
learned neural network that guides program
search~\cite{spiral,balog2016deepcoder,devlin2017neural}, while recent
symbolic AI research has developed frameworks for learning the DSL by
inferring reusable pieces of
code~\cite{ecc,Dechter:2013:BLV:2540128.2540316,DBLP:conf/icml/LiangJK10,DBLP:conf/ecai/LinDETM14}.
A main goal of this work is to combine and refine these techniques
with the intention of building agents that, like humans, develop
domain expertise for new classes of problems.  Work in cognitive
modeling has often exploited program-like representations for both
concept~\cite{piantadosi2011learning,lake2015human,GoodmanEtAl2015-Chapter} and intuitive theory
learning~\cite{logical,Ullman2012},
and we view these models of concept learning as
a kind of program induction,
and see theory learning as similar to inducing a DSL or library.

Deep learning is a dominant machine intelligence paradigm that
has had many successes
at creating machines with impressive expertise at
specific tasks in vision, natural language, gameplaying, and many other areas. ideas: we want to become an expert from modest amounts of data; we want to be able to transfer to different types of problems, and to do this zero shot;

@@

    %% center spiral
    \begin{scope}[shift={(3.25,-7.9)},scale=0.8]    
      \spiral{(3.5,1)}{3.5}
      \node[latent,scale=1] at (3.5,3) (dx){Library};
      \node[latent,scale=1] at ([yshift=-1.7cm,xshift=0cm]dx) (zp){prog};
      \node[obs,scale=1] at ([yshift=-1.45cm]zp) (xp) {task};
      \node[latent,scale=1] at ([xshift=2cm]zp) (zp1){prog};
      \node[obs,scale=1] at ([xshift=2cm]xp) (xp1) {task};
      \draw [->] (zp1.south) -- (xp1.north);
      \draw [->] (dx.south) -- (zp1.north);
      \draw [->,red] (xp1.east) to[out = 30,in = -30] node(nn){} (zp1.east);

@@

  \end{tikzpicture}
  \caption{\textbf{Middle:} \system as a graphical model. Agent observes programming tasks (e.g., input/outputs for list processing or images for graphics programs), which it explains with latent programs, while jointly inferring a latent Domain Specific Language (DSL) capturing cross-program regularities. A neural network, called the \emph{recognition model} (red arrows) is trained to quickly infer programs with high posterior probability. \textbf{Top}: Wake phase infers programs while holding the DSL and recognition model fixed. \textbf{Left}: Sleep (Consolidation) phase infers DSL while holding the programs fixed by refactoring programs found during waking and extracting common components. \textbf{Right}: Sleep (Dreaming) phase trains recognition model to predict approximate posterior over programs conditioned on task. Trained on `Fantasies' (programs sampled from DSL) \& `Replays' (programs found during waking).}\label{threeCycles}
\end{figure}

@@

\subsection{Consolidation-Sleep: Growing a Domain Specific Language}\label{consolidationSection}

The DSL offers a set of abstractions that allow an agent to concisely
express solutions to the tasks at hand. We automatically discover
these new abstractions by combining two ideas. First, we build on
techniques from the programming languages community to develop a new
algorithm for automatically refractoring programs, where this
refactoring exposes common reused subexpressions across the
programs found during waking.  Second, we use this automatic refactoring
process to search for DSLs that maximally compress these programs by
incorporating reused subexpressions into the DSL.

Mathematically this compression takes the form of
finding the DSL maximizing $\int \probability[\mathcal{D},\theta]\probability[X|\mathcal{D},\theta]\;\mathrm{d}\theta$ (Sec.~\ref{overviewSection}).
We replace this marginal with an AIC approximation~\cite{akaike1998information}
and marginalize over refactorings of programs found during waking, minimizing the following expression,
which can be interpreted as a kind of compression:
\begin{equation}
\underbrace{-\log \probability[\mathcal{D}] + \min_{\substack{\theta\\\\\\\\}}\Bigg(-\log
\probability[\theta|\mathcal{D}] + \|\theta\|_0}_{\text{Description length of }(\mathcal{D},\theta)} +\sum_{x\in
  X}\underbrace{-\log \sum_{\substack{p \text{ a refactoring of } p'\\ p'\text{ found during waking}}}\probability[x|p]\probability[p|\mathcal{D},\theta]}_{\text{Description length of programs for task }x}\Bigg)
\label{AIC}
\end{equation}
But a program has infinitely many possible refactorings, rendering
Eq.~\ref{AIC} intractable.  Rather than consider every refactoring we
bound the number of $\lambda$-calculus evaluation steps separating a
program from its refactoring.  Now the number of refactorings is
finite but astronomically large: Figure~\ref{mapFactor}A diagrams a
problem where the agent rediscovers the higher-order function
\code{map} starting from the basics of Lisp and the Y-combinator, but
where there are approximately $10^{14}$ possible refactorings -- a quantity
that grows exponentially both as a function of program size and a
function of the bound on evaluation steps. How can we tame this
combinatorial explosion?

To resolve this exponential growth we introduce a new data structure
combining ideas from version space
algebras~\cite{lau2001programming,mitchell1977version,polozov2015flashmeta}
and equivalence graphs~\cite{tate2009equality}. A version space is a
tree-shaped data structure that compactly represents a large set of
programs and supports efficient set operations like union,
intersection, and membership checking, while equivalence graphs are
data structures that track semantic equivalences between program
subexpressions.  In Appendix~\ref{appendixVersion}, we give a dynamic
program that takes as input a program and then outputs a version space
containing its refactorings while tracking semantically equivalent
subexpressions. Figure~\ref{mapFactor}B diagrams a subtree of a
version space containing refactorings of a small program. Our
technique is substantially more efficient than explicitly representing
the space of possible refactorings: for the example in
Figure~\ref{mapFactor}A, we represent the space of refactorings using
a version space with $10^6$ nodes, which encodes $10^{14}$
refactorings. Appendix~\ref{appendixCompression} specifies how we
combine this probabilistic and symbolic machinery to update the DSL.
At a high level, our approach is to search locally through the space
of DSLs, proposing small changes until Eq.~\ref{AIC} fails to
decrease.

@@

  \end{tikzpicture}
  \caption{DSL learning as code refactoring. \textbf{Panel A:} During waking we discover programs for each task, then refactor the code from those programs to expose common subprograms (highlighted in \orange{orange}). Common subprograms are incorporated into the DSL when they increase a Bayesian objective. Intuitively, these new DSL components best compress the programs found during waking. \textbf{Panel B:} \# of possible refactorings grows exponentially with program size, so we represent refactorings using version spaces, which augment syntax trees with a \emph{union} operator whose children are themselves version spaces. Right graph: version spaces are exponentially more efficient than explicitly constructing set of refactorings. In this graph, refactored programs are of the form $1+1+\cdots  + 1$.}\label{mapFactor}
\end{figure*}

@@

During ``dreaming'' the system learns a recognition model that guides
program search.  It learns from (program, task) pairs drawn from two
sources of self-supervised data: \emph{replays} of programs discovered
during waking, and \emph{fantasies}, or programs drawn from the DSL.
Replays ensure that the recognition model is trained on the actual
tasks it needs to solve, and does not forget how to solve them.
Fantasies ensure that the recognition model has a large and highly
varied corpus of (program, task) pairs to learn from.

@@

\noindent\textbf{Parameterizing $Q$.} The recognition model predicts a
fixed-dimensional tensor encoding a distribution over subroutines in
the DSL, conditioned on the local context in the syntax tree of the
program. This local context consists of the parent node in the syntax
tree, as well as which argument is being generated, functioning as a
kind of `bigram' model over trees. Figure~\ref{symmetry} (left)
diagrams this generative process.  This parameterization confers three
main advantages: (1) it supports fast enumeration and sampling of
programs, because the recognition model only runs once per
task, like in~\cite{balog2016deepcoder,ecc,menon2013machine} -- thus
we can fall back on fast enumeration if the target program is unlike
the training programs; (2) the recognition model provides fine-grained
information about the structure of the target program, similar
to~\cite{devlin2017robustfill,zavershynskyi2018naps}; and (3) in
conjunction with $\mathcal{L}^{\text{MAP}}$ the recognition model
learns to break symmetries in the space of programs.

\noindent\textbf{Symmetry breaking.} A good DSL not only exposes high-level building blocks, but also
carefully restricts the ways in which those building blocks are
allowed to compose.  For example,
%% a DSL for list manipulation should
%% contain both the empty list and a routine for appending lists, but
%% should not allow appending the empty list.  Similarly 
a DSL for arithmetic should disallow adding zero, or force
right-associative addition. A bigram parameterization of the recognition
model, combined with the $\mathcal{L}^{\text{MAP}}$ training
objective, interact in a way that breaks symmetries like these, allowing the agent to more efficiently explore the space of
programs.  This interaction occurs because the bigram parameterization
can disallow DSL primitives depending on their local syntactic
context, while the $\mathcal{L}^{\text{MAP}}$ objective forces all
probability mass onto a single member of a set of syntactically
distinct but semantically equivalent expressions
(Appendix~\ref{recognitionAppendix}).
We experimentally confirm this symmetry-breaking
behavior by training recognition models that minimize either
$\mathcal{L}^{\text{MAP}}$/$\mathcal{L}^\text{posterior}$ and which
use either a bigram parameterization/unigram\footnote{In the unigram variant $Q$ predicts a $|\mathcal{D}| + 1$-dimensional vector: $Q(p|x) = \probability[p|\mathcal{D},\theta_i = Q_i(x)]$,
  and was used in our prior work~\cite{ecc}} parameterization.
Figure~\ref{symmetry} (right) shows the result of training $Q$ in these four regimes
and then sampling programs.
On this particular run,
the combination of
bigrams and $\mathcal{L}^{\text{MAP}}$ learns to
avoid adding zero and associate addition to the right ---
different random initializations
lead to either right or left association.

@@

\subsubsection{List Processing}\label{listSection}
We took 218 list manipulation tasks from our previous work~\cite{ecc},
each with 15 input/output examples.  In solving these tasks, the
system typically composes around 20 new DSL components, and discovered multiple
higher-order functions. Each round of memory consolidation built on
components discovered in earlier sleep cycles --- for example the
agent first learns the higher-order function \code{filter}, uses
\code{filter} to learn to take the maximum element of a list, then
uses that routine to learn a new component for extracting the
$n^{\text{th}}$ largest element of a list, which it finally uses to
solve a task involving sorting a list of numbers (Figure~\ref{initialExampleDSL}).
This incremental, modular learning of deep hierarchies of DSL
components occurs because of the alternation between code writing
(during waking) and code refactoring (during the consolidation phase
of sleep).

%% \begin{figure}[b]\centering
%% \vspace{-0.5cm}  \begin{tabular}{lll}
%%     \toprule
%%     Name & Input & Output \\\midrule
%%     repeat-3 & [7\, 0] & [7\, 0\, 7\, 0\, 7\, 0] \\
%%     drop-3 & [0\, 3\, 8\, 6\, 4] & [6\, 4] \\
%%     rotate-2 & [8\, 14\, 1\, 9] & [1\, 9\, 8\, 14] \\
%%     count-head-in-tail & [1\, 2\, 1\, 1\, 3] & 2 \\
%%     keep-div-5 & [5\, 9\, 14\, 6\, 3\, 0] & [5\, 0] \\
%%     product & [7\, 1\, 6\, 2] & 84 \\
%%     \bottomrule
%%   \end{tabular}
%%   \captionof{table}{Some tasks in our list function domain.}\label{listExamples}\vspace{-0.5cm}
%% \end{figure}

\subsubsection{Text Editing}\label{textSection}
Synthesizing programs that edit text is a classic problem in the
programming languages and AI literatures~\cite{lau2001programming},
and algorithms that synthesize text editing programs ship in Microsoft
Excel~\cite{gulwani2011automatin}.  This prior work uses
hand-engineered DSLs and hand-engineered search strategies.  Here, we
will show that we can jointly learn both these ingredients and surpass
the state-of-the-art domain-general program synthesizers on a standard
text editing benchmark.

@@

Procedural visual concepts are studied across AI and cognitive science
--- Bongard problems~\cite{Moscow}, Raven's progressive
matrices~\cite{raven2003raven}, and Lake et al.'s BPL model of
omniglot~\cite{lake2015human} are prominent examples.  Here we take
inspiration from LOGO Turtle graphics~\cite{turtle}, tasking our agent
with drawing a modest corpus of images while equipping it with control
over a `pen', along with arithmetic operations on angles and
distances.

Inside its learned DSL we find interpretable parametric drawing
routines corresponding to the families of visual objects in its
training data, like polygons, circles, and spirals
(Figure~\ref{logoPrimitives}, left) -- without supervision the agent
has learned the basic types of objects in its visual world. It
additionally learns more abstract visual relationships, like
rotational symmetry, which it models by incorporating a higher-order
function into its DSL (Figure~\ref{logoPrimitives}, right).  %% This
%% abstraction comes from jointly compressing programs for all of the
%% training images into its DSL.

What does \system dream of?  Prior to learning samples from the DSL
are simple and largely unstructured (Figure~\ref{logoDreams}, left).
After training the samples become richly structured
(Figure~\ref{logoDreams}, right), compositionally recombining latent
building blocks and motifs acquired from the training data. This
offers a visual window into how the generative model bootstraps recognition
model training: as the DSL grows more finely tuned to the domain, the
neural net gets richer and more highly varied training data.

@@

Inspired by the classic AI `copy task' --- where an agent must look at
an image of a tower made of toy blocks and re-create the
tower~\cite{towerCopy} --- we give \system 112 tower `copy tasks'
(Figure~\ref{towerTasks}).  Here the agent observes both an image of a
tower and the locations of each of its blocks, and must write a
program that plans how a simulated hand would build the tower.
These towers are built from Lego-style blocks
that snap together on a discrete grid.
The system starts with the same control flow primitives as with
LOGO graphics,
and learns parametric `options' for building blocks towers (Figure~\ref{tower}),
inferring concepts like arches, staircases, bridges and brick walls.
\begin{figure}
  \begin{tabular}{cc}
    \toprule Example learned DSL components&Samples from DSL\\\midrule
    \begin{tabular}{rl}
      Brickwall&\begin{tabular}{l}
        \includegraphics[width = 5cm]{figures/tower/tower_dsl_bricks.png}
      \end{tabular}\\
      Bridge&\begin{tabular}{l}
        \includegraphics[width = 5cm]{figures/tower/tower_dsl_bridge.png}
      \end{tabular}\\
      Staircase&\begin{tabular}{l}
        \includegraphics[width = 5cm]{figures/tower/tower_dsl_staircase.png}
        \end{tabular}
    \end{tabular}&
    \begin{tabular}{l}
      \includegraphics[width = 5cm]{figures/tower/dreams.png}
      \end{tabular}
    \\\bottomrule 
    \end{tabular}
  \caption{\textbf{Left}: Three (out of 19) learned DSL components for building towers out of Lego-style blocks. These components act like parametric options~\cite{stolle2002learning},
    giving higher-level building blocks that the agent can use to plan. \textbf{Right}: 16 random samples, or `dreams', from learned DSL.}\label{tower}
\end{figure}

@@

\subsubsection{Generative Modeling of Text}
We investigate few-shot learning of generative models by tasking our
agent with inferring a probabilistic regular expression from a small
number (5) of strings - for example, observing the strings \$1.20 and \$9.42
and predicting the regex \$d

For this task, we used the dataset from [cite Luke], consisting of a
large corpus of string concepts. Each string concept has 5 ``train"
strings and 5 ``test" strings.

We trained \system using 128 string concepts, randomly sampled from
the corpus, and tested on another 128 randomly sampled string
concepts.

Our initial DSL consisted of all printable characters, four special
character classes (lowercase letters $\\l$, uppercase letters $\\u$,
digits $\\d$, and '.', which represents all possible characters), and
3 functions: \textbf{Kleene Star}, \textbf{Maybe}, \textbf{Or}. We
also included a task-dependent 'constant' token, which prints the
largest constant string shared across the 5 training tasks.
Concatenation was done via continuation-style, as above in the logo
domain ... [idk how you wanted this to be mentioned?]

Because this domain has a relatively large number of parameters, we
used a modified version of the recognition model, where only unigram
parameters were inferred. We found that this recognition model trained
quicker and more had more stable convergence.

Probabilistic regexes have no ground truth notion of ``solved," so we
evaluate differently from other domains. Our main evaluation metric is
to report a lower bound on the posterior predictive of each task given
the learned grammar:

>>>>

\begin{figure}
  \includegraphics[width = \textwidth]{figures/dpl/list}
  \includegraphics[width = \textwidth]{figures/DeepLogo_4_11}
  \caption{}\label{exampleDSL}
\end{figure}

@@

Our model, called \systemEnding, works through a novel kind
of wake/sleep or `dream' learning~\cite{hinton1995wake}, iterating
through a wake cycle -- where it solves \textbf{tasks} by writing programs (Figure~\ref{threeCycles} top)
-- and a pair of sleep cycles, both of which are loosely biologically
inspired. The first sleep cycle, which we refer to as
\textbf{abstraction}, grows its library of code by replying
experiences from waking and consolidating them into new code
abstractions (Figure~\ref{threeCycles} left).  This cycle is inspired by the formation of abstractions
during sleep memory consolidation~\cite{DUDAI201520}.  The second
sleep cycle, which we refer to as \textbf{dreaming}, improves the
agents knowledge of how to write code by training a neural network to
help quickly search for programs. The neural net is trained on replayed
experiences as well as `fantasies', or samples, from the learned
library (Figure~\ref{threeCycles} right).  These two kinds of dreams are inspired by the distinct
episodic replay and hallucination components of dream
sleep~\cite{fosse2003dreaming}.

@@

\begin{array}{lr}
  \begin{array}{l}  p_x = \argmax_p \probability[x|p]\probability[p|\mathcal{D}^*,\theta^*]\text{, for each task $x\in X$}\end{array}
    &\text{\emph{Wake}}\\\\
\begin{array}{rl}  
\mathcal{D}^* = \argmax_\mathcal{D}&\int\probability\left[\mathcal{D},\phantom{^*}\theta \right]\prod_{x\in X}\sum_{\substack{p\text{ found during waking}}} \probability[x|p]\probability[p|\mathcal{D},\phantom{^*}\theta]\;\text{d}\theta\\
\theta^* = \argmax_\theta&\phantom{\int}\probability\left[\mathcal{D}^*,\theta \right]\prod_{x\in X}\sum_{p\text{ found during waking}} \probability[x|p]\probability[p|\mathcal{D}^*,\theta]
 \end{array}
&\text{\emph{Abstraction (sleep)}}\\\\
\begin{array}{l} Q(p|x)\approx \probability[p|x,\mathcal{D}^*,\theta^*]\end{array}
  &\text{\emph{Dreaming (sleep)}}
\end{array}

@@

@@

    %% center spiral
    \begin{scope}[shift={(3.25,-7.9)},scale=0.8]    
      \spiral{(3.5,1)}{3.5}
      \node[latent,scale=1] at (3.5,3) (dx){Library};
      \node[latent,scale=1] at ([yshift=-2cm,xshift=0cm]dx) (zp){prog};
      \node[obs,scale=1] at ([yshift=-1.45cm]zp) (xp) {task};
      \node[latent,scale=1] at ([xshift=2cm]zp) (zp1){prog};
      \node[obs,scale=1] at ([xshift=2cm]xp) (xp1) {task};
      \draw [->] (zp1.south) -- (xp1.north);
      \draw [->] (dx.south) -- (zp1.north);
      \draw [->,red] (xp1.east) to[out = 30,in = -30] node(nn){} (zp1.east);

@@

  \end{tikzpicture}
  \caption{\textbf{Middle:} \system as a graphical model. Agent observes programming tasks (e.g., input/outputs for list processing or images for graphics programs), which it explains with latent programs, while jointly inferring a latent library capturing cross-program regularities. A neural network, called the \emph{recognition model} (red arrows) is trained to quickly infer programs with high posterior probability. \textbf{Top}: Wake phase infers programs while holding the library and recognition model fixed. \textbf{Left}: Sleep (Consolidation) phase updates library while holding the programs fixed by refactoring programs found during waking and extracting common components. \textbf{Right}: Sleep (Dreaming) phase trains recognition model to predict approximate posterior over programs conditioned on task. Trained on `Fantasies' (programs sampled from library) \& `Replays' (programs found during waking).}\label{threeCycles}
\end{figure}

@@

\subsection{Abstraction Sleep: Growing a Library of Concepts}\label{consolidationSection}

The library offers a set of abstractions that allow an agent to concisely
express solutions to the tasks at hand. We automatically discover
these new abstractions by combining two ideas. First, we build on
techniques from the programming languages community to develop a new
algorithm for automatically refractoring programs, where this
refactoring exposes common reused subexpressions across the
programs found during waking.  Second, we use this automatic refactoring
process to search for libraries that maximally compress these programs by
incorporating reused subexpressions into the DSL.

Mathematically this compression takes the form of
finding the library maximizing $\int \probability[\mathcal{D},\theta]\probability[X|\mathcal{D},\theta]\;\mathrm{d}\theta$ (Sec.~\ref{overviewSection}).
We replace this marginal with an AIC approximation~\cite{akaike1998information}
and marginalize over refactorings of programs found during waking, minimizing the following expression,
which can be interpreted as a kind of compression:
\begin{equation}
\underbrace{-\log \probability[\mathcal{D}] + \min_{\substack{\theta\\\\\\\\}}\Bigg(-\log
\probability[\theta|\mathcal{D}] + \|\theta\|_0}_{\text{Description length of }(\mathcal{D},\theta)} +\sum_{x\in
  X}\underbrace{-\log \sum_{\substack{p \text{ a refactoring of } p'\\ p'\text{ found during waking}}}\probability[x|p]\probability[p|\mathcal{D},\theta]}_{\text{Description length of programs for task }x}\Bigg)
\label{AIC}
\end{equation}
But a program has infinitely many possible refactorings, rendering
Eq.~\ref{AIC} intractable.  Rather than consider every refactoring we
bound the number of $\lambda$-calculus evaluation steps separating a
program from its refactoring.  Now the number of refactorings is
finite but astronomically large: Figure~\ref{mapFactor}A diagrams a
problem where the agent rediscovers the higher-order function
\code{map} starting from the basics of Lisp and the Y-combinator, but
where there are approximately $10^{14}$ possible refactorings -- a quantity
that grows exponentially both as a function of program size and a
function of the bound on evaluation steps. How can we tame this
combinatorial explosion?

To resolve this exponential growth we introduce a new data structure
combining ideas from version space
algebras~\cite{lau2001programming,mitchell1977version,polozov2015flashmeta}
and equivalence graphs~\cite{tate2009equality}. A version space is a
tree-shaped data structure that compactly represents a large set of
programs and supports efficient set operations like union,
intersection, and membership checking, while equivalence graphs are
data structures that track semantic equivalences between program
subexpressions.  In Appendix~\ref{appendixVersion}, we give a dynamic
program that takes as input a program and then outputs a version space
containing its refactorings while tracking semantically equivalent
subexpressions. Figure~\ref{mapFactor}B diagrams a subtree of a
version space containing refactorings of a small program. Our
technique is substantially more efficient than explicitly representing
the space of possible refactorings: for the example in
Figure~\ref{mapFactor}A, we represent the space of refactorings using
a version space with $10^6$ nodes, which encodes $10^{14}$
refactorings. Appendix~\ref{appendixCompression} specifies how we
combine this probabilistic and symbolic machinery to update the library.
At a high level, our approach is to search locally through the space
of libraries, proposing small changes until Eq.~\ref{AIC} fails to
decrease.

@@

  \end{tikzpicture}
  \caption{Library learning as code refactoring. \textbf{Panel A:} During waking we discover programs for each task, then refactor the code from those programs to expose common subprograms (highlighted in \orange{orange}). Common subprograms are incorporated into the library when they increase a Bayesian objective. Intuitively, these new components best compress the programs found during waking. \textbf{Panel B:} \# of possible refactorings grows exponentially with program size, so we represent refactorings using version spaces, which augment syntax trees with a \emph{union} operator whose children are themselves version spaces. Right graph: version spaces are exponentially more efficient than explicitly constructing set of refactorings. In this graph, refactored programs are of the form $1+1+\cdots  + 1$.}\label{mapFactor}
\end{figure*}

@@

During ``dreaming'' the system learns a recognition model that guides
program search.  It learns from (program, task) pairs drawn from two
sources of self-supervised data: \emph{replays} of programs discovered
during waking, and \emph{fantasies}, or programs drawn from $(\mathcal{D},\theta)$.
Replays ensure that the recognition model is trained on the actual
tasks it needs to solve, and does not forget how to solve them.
Fantasies ensure that the recognition model has a large and highly
varied corpus of (program, task) pairs to learn from.

@@

\noindent\textbf{Parameterizing $Q$.} The recognition model predicts a
fixed-dimensional tensor encoding a distribution over routines in
the library, conditioned on the local context in the syntax tree of the
program. This local context consists of the parent node in the syntax
tree, as well as which argument is being generated, functioning as a
kind of `bigram' model over trees. Figure~\ref{symmetry} (left)
diagrams this generative process.  This parameterization confers three
main advantages: (1) it supports fast enumeration and sampling of
programs, because the recognition model only runs once per
task, like in~\cite{balog2016deepcoder,ecc,menon2013machine} -- thus
we can fall back on fast enumeration if the target program is unlike
the training programs; (2) the recognition model provides fine-grained
information about the structure of the target program, similar
to~\cite{devlin2017robustfill,zavershynskyi2018naps}; and (3) in
conjunction with $\mathcal{L}^{\text{MAP}}$ the recognition model
learns to break symmetries in the space of programs.

\noindent\textbf{Symmetry breaking.} Effective domain-specific representations not only exposes high-level building blocks, but also
carefully restrict the ways in which those building blocks are
allowed to compose.  For example,
%% a DSL for list manipulation should
%% contain both the empty list and a routine for appending lists, but
%% should not allow appending the empty list.  Similarly 
when searching over arithmetic expressions, one could disallow adding zero, and force
right-associative addition. A bigram parameterization of the recognition
model, combined with the $\mathcal{L}^{\text{MAP}}$ training
objective, interact in a way that breaks symmetries like these, allowing the agent to more efficiently explore the space of
programs.  This interaction occurs because the bigram parameterization
can disallow library routines depending on their local syntactic
context, while the $\mathcal{L}^{\text{MAP}}$ objective forces all
probability mass onto a single member of a set of syntactically
distinct but semantically equivalent expressions
(Appendix~\ref{recognitionAppendix}).
We experimentally confirm this symmetry-breaking
behavior by training recognition models that minimize either
$\mathcal{L}^{\text{MAP}}$/$\mathcal{L}^\text{posterior}$ and which
use either a bigram parameterization/unigram\footnote{In the unigram variant $Q$ predicts a $|\mathcal{D}| + 1$-dimensional vector: $Q(p|x) = \probability[p|\mathcal{D},\theta_i = Q_i(x)]$,
  and was used in our prior work~\cite{ecc}} parameterization.
Figure~\ref{symmetry} (right) shows the result of training $Q$ in these four regimes
and then sampling programs.
On this particular run,
the combination of
bigrams and $\mathcal{L}^{\text{MAP}}$ learns to
avoid adding zero and associate addition to the right ---
different random initializations
lead to either right or left association.

@@

\subsubsection{List Processing}\label{listSection}
We took 218 list manipulation tasks from our previous work~\cite{ecc},
each with 15 input/output examples.  In solving these tasks, the
system typically composes around 20 new library routines, and discovered multiple
higher-order functions. Each round of memory consolidation built on
components discovered in earlier sleep cycles --- for example the
agent first learns the higher-order function \code{filter}, uses
\code{filter} to learn to take the maximum element of a list, then
uses that routine to learn a new component for extracting the
$n^{\text{th}}$ largest element of a list, which it finally uses to
solve a task involving sorting a list of numbers (Figure~\ref{initialExampleDSL}).
This incremental, modular learning of deep hierarchies of library
components occurs because of the alternation between code writing
(during waking) and code refactoring (during the abstraction phase
of sleep).

\subsubsection{Text Editing}\label{textSection}
Synthesizing programs that edit text is a classic problem in the
programming languages and AI literatures~\cite{lau2001programming},
and algorithms that synthesize text editing programs ship in Microsoft
Excel~\cite{gulwani2011automatin}.  This prior work uses
hand-engineered libraries of primitives and hand-engineered search strategies.  Here, we
will show that we can jointly learn both these ingredients and surpass
the state-of-the-art domain-general program synthesizers on a standard
text editing benchmark.

@@

Procedural visual concepts are studied across AI and cognitive science
--- Bongard problems~\cite{Moscow}, Raven's progressive
matrices~\cite{raven2003raven}, and Lake et al.'s BPL model of
omniglot~\cite{lake2015human} are prominent examples.  Here we take
inspiration from LOGO Turtle graphics~\cite{turtle}, tasking our agent
with drawing a corpus of 160 images while equipping it with control
over a `pen', along with arithmetic operations on angles and
distances.

Inside its learned library we find interpretable parametric drawing
routines corresponding to the families of visual objects in its
training data, like polygons, circles, and spirals
(Figure~\ref{logoPrimitives}, left) -- without supervision the agent
has learned the basic types of objects in its visual world. It
additionally learns more abstract visual relationships, like
rotational symmetry, which it models by abstracting out a new higher-order
function into its library (Figure~\ref{logoPrimitives}, right).

What does \system dream of?  Prior to learning random programs written using the library
are simple and largely unstructured (Figure~\ref{logoDreams}, left).
After training the samples become richly structured
(Figure~\ref{logoDreams}, right), compositionally recombining latent
building blocks and motifs acquired from the training data. This
offers a visual window into how the generative model bootstraps recognition
model training: as the library grows more finely tuned to the domain, the
neural net gets richer and more highly varied training data.

@@

Inspired by the classic AI `copy task' --- where an agent must look at
an image of a tower made of toy blocks and re-create the
tower~\cite{towerCopy} --- we give \system 112 tower `copy tasks'
(Figure~\ref{towerTasks}).  Here the agent observes both an image of a
tower and the locations of each of its blocks, and must write a
program that plans how a simulated hand would build the tower.
These towers are built from Lego-style blocks
that snap together on a discrete grid.
The system starts with the same control flow primitives as with
LOGO graphics,
and learns parametric `options' for building blocks towers (Figure~\ref{tower}),
inferring concepts like arches, staircases, bridges and brick walls.
\begin{figure}
  \begin{tabular}{cc}
    \toprule Example learned DSL components&Samples from DSL\\\midrule
    \begin{tabular}{rl}
      Brickwall&\begin{tabular}{l}
        \includegraphics[width = 5cm]{figures/tower/tower_dsl_bricks.png}
      \end{tabular}\\
      Bridge&\begin{tabular}{l}
        \includegraphics[width = 5cm]{figures/tower/tower_dsl_bridge.png}
      \end{tabular}\\
      Staircase&\begin{tabular}{l}
        \includegraphics[width = 5cm]{figures/tower/tower_dsl_staircase.png}
        \end{tabular}
    \end{tabular}&
    \begin{tabular}{l}
      \includegraphics[width = 5cm]{figures/tower/dreams.png}
      \end{tabular}
    \\\bottomrule 
    \end{tabular}
  \caption{\textbf{Left}: Three (out of 19) learned DSL components for building towers out of Lego-style blocks. These components act like parametric options~\cite{stolle2002learning},
    giving higher-level building blocks that the agent can use to plan. \textbf{Right}: 16 random samples, or `dreams', built from learned library.}\label{tower}
\end{figure}

@@

\subsubsection{Generative Modeling of Text}
Few-shot learning of simple generative concepts comes naturally to
humans, from learning new rules in natural language~\cite{marcus1999rule},
to learning routines for symbols and signs
\cite{lake2015human},
to learning new motor routines for producing words.
We investigate few-shot generative modeling by tasking our
agent with inferring a probabilistic regular expression from a small
number of strings, where these strings are drawn from CSV columns crawled from the web~\cite{LUKE}. Across wake/sleep cycles, our agent learns to learn
regular expressions that describe the structure of a typically occurring text concepts,
like numbers or dates.
Figure~\ref{qualitativeRegex} shows example model outputs on held out generative modeling tasks,
contrasting the full model with ablations that lesion either the library learning or recognition model training.
\begin{figure}
  \includegraphics[width = \textwidth]{figures/regexQualitative.pdf}
  \caption{Qualitative results on probabilistic generative text modeling. Agent observes 5 strings (`Input' column) and infers a probabilistic regex (MAP program--regex primitives highlighted in red), from which it can imagine more examples of the latent text concept. No Library: Dreaming only (ablates library learning). No Rec: Consolidation only (ablates recognition model)}\label{qualitativeRegex}
\end{figure}

\subsection{From libraries to languages}

Rather than start with a domain-specific basis and then enrich it over successive wake/sleep cycles -- as these prior experiments have done --
could we instead start with a highly generic basis,
and then learn a library encoding something closer to the basic domain language?
We first consider learning a basic language for equations describing physical laws,
starting \emph{without} concepts like vectors or forces
and instead providing sequences of real numbers and Lisp-style sequence manipulation.
We collected 57 equations for physical laws from AP and MCAT physics `cheat sheets' --
forms like Coulomb's law and expressions for the center of mass of a set of objects --
along with
reference vector algebra identities from machine learning textbooks,
and tasked our agent with inferring programs calculating these input/output mappings.
In solving these tasks, \system first discovers the building blocks of vector algebra,
like inner products, vector sums, and norms,
and then uses this vocabulary to infer library routines for common
physical relationships like inverse square laws and integrating second derivatives over time.
